---
title: ' Machine Learning | Prediction Assignment'
author: "Kimamö Wachira"
date: "August 27, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 setwd("~/Google Drive/Coursera/Data Science Specialization/Machine Learning/Predictive Assignment")

```{r} 

library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(e1071)
library(lubridate)
library(RColorBrewer) 

```

## Introduction


## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).
 
```{r get data, cache=TRUE}

## Download and read the training data
trainingFile <-  "pml-training.csv"
if (!file.exists(trainingFile)) {
    url <-
    "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
    trainingFile <- download.file(url, destfile = trainingFile)
  }

trainingData <-
    read.csv(trainingFile, na.strings = c("NA", "#DIV/0!", ""))

## Download and read the testing data
testingFile <-  "pml-testing.csv"
if (!file.exists(testingFile)) {
    url <-
    "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
    testingFile <- download.file(url, destfile = testingFile)
  }

testingData <-
    read.csv(testingFile, na.strings = c("NA", "#DIV/0!", ""))

```

## Data Exploration and Cleaning
###### use variables in testingData only and disregard the other columns in trainingData set that not needed and exclude NA and Empty columns
```{r 'data cleaning'  }

testVariables <-
  names(testingData[, colSums(is.na(testingData)) == 0])[8:59]
  head(testVariables)
  
  
# exclude NA and Empty columns
  trainingData <-
    trainingData[, names(trainingData)[sapply(trainingData, function(x)
    ! (any(is.na(x) | x == "")))]]
  
# set the variables necessary to use as per test dataset
  trainingData <- trainingData[, c(testVariables, "classe")]
  testingData <- testingData[, c(testVariables, "problem_id")]
  
```

## create DataPartition
###### 60% data goes to training set and the rest to testing set
```{r, createDataPartition}
set.seed(12345)

# Divide the trainingData into training set & testing set
inTrain <- createDataPartition(trainingData$classe, p=0.6, list=FALSE)
training <- trainingData[inTrain,]
testing <- trainingData[-inTrain,]

dim(training); 
dim(testing);

```
## Build a Model
#### using Random Forest to build a predictive model
```{r, PredictiveModel, cache=TRUE}

fitControl <-  trainControl(method = "cv", 5)

rfModel <- rpart(classe ~ ., data = training, method = "class")
      #     train(
      #     classe ~ .,
      #     data = training,
      #     method = "rf",
      #     preProcess = c("center", "scale"),
      #     trControl = fitControl,
      #     ntree = 250
      # )

#rfModel 

## Plot model
plot(rfModel)

```
#### Predict the Model: Measure Model accuracy and Sample Error 
```{r}

set.seed(10)
 trainingPrediction <- predict(rfModel, newdata = training, method="class")
 print(confusionMatrix(trainingPrediction, training$classe ))

```
##### Predict model: Measure testing data accuracy and sample error

```{r testing accuracy}

  testPredication <- predict(rfModel, newdata= testing, method="class")
  print(confusionMatrix(testPredication, testing$classe ))
 
```

## create a boost model

```{r rfFitModel }

  set.seed(12345)
  rfFitModel <- randomForest(classe ~ ., data = training, ntree = 1000)
#     boostModel <- 
#               train(
#               classe ~ .,
#               data = training,
#               method = "rf",
#               verbose = F,
#               preProcess = c("center", "scale"),
#               trControl = fitControl,
#               ntree = 250
#           )
# 
# boostModel
# plot(boostModel)


## track the changes in rfModel using varImp function 
  varImp(rfFitModel)

  ##rfModel$finalModel
```

```{r apply the prediction model, cache=TRUE}

  modelPrediction <- predict(rfFitModel, testing, type="class")
  testPred <- cbind(modelPrediction,testing)

  ## Boost Predict
    # modelPrediction <- predict(boostModel, testing)
    # print(confusionMatrix(modelPrediction, testing$classe ))
    # 
    # 
  
  subset(testing, select=names(testing)[grep("belt|[^(fore)]arm|dumbbell|forearm", names(testing), invert = TRUE)])
  
  print(rfFitModel)

  
## write to a file for Assignement Quiz answers  
createQuizAnswer = function(x){
  n = length(x)
  
  #AssignmentDir <- "Predictive Assignment"
  QuizAnswerDir <- "Quiz-Answers"
  
 # path <- file.path(AssignmentDir,QuizAnswerDir)
  
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=file.path(QuizAnswerDir,filename),quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
createQuizAnswer(modelPrediction)
     

```

