---
title: "Machine Learning | Predictive Assignment Writeup"
author: "Kimamö Wachira"
date: "10/3/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
 setwd("~/Google Drive/Coursera/Data Science Specialization/Machine Learning/Predictive Assignment")

 ## libraries that might require
 library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(e1071)
library(lubridate) 
library(RColorBrewer) 
library(Hmisc)
library(plyr)
```
## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. [More information is available from the website here:] (http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

### Data

[The training data for this project are available here:] (https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)

[The test data are available here:] (https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

[The data for this project come from this source:] (http://groupware.les.inf.puc-rio.br/har). If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

### Goal

The goal of your project is to predict the manner in which they did the exercise. This is the “classe” variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

### Fetch Data

```{r datafetch, cache=TRUE}
 
## Download and read the training data
training.csv <-  "pml-training.csv"
if (!file.exists(training.csv)) {
    url <-
    "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
    training.csv <- download.file(url, destfile = training.csv)
}

## Download and read the testing data
testing.csv <-  "pml-testing.csv"
if (!file.exists(testing.csv)) {
    url <-
    "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
    testing.csv <- download.file(url, destfile = testing.csv)
  }


## Load data to memory
training <-
    read.csv(training.csv, na.strings = c("NA", "#DIV/0!", ""),stringsAsFactors = FALSE)

testing <-
    read.csv(testing.csv, na.strings = c("NA", "#DIV/0!", ""),stringsAsFactors = FALSE)

dim(training)  
 

```

## clean-up the data

```{r cleanUp}

training$classe <- as.factor(training$classe)
training.clean <- training[,-nearZeroVar(training)]

### remove unnecesary 7 columns
training.clean <- training.clean[,-c(1,2,3,4,5,6,7)]

uIndex <- sapply(training.clean, function(i) i == "#DIV/0!")
training.clean[uIndex] <- NA

## after some initial cleaning
  dim(training.clean)

testVariables <-
  names(testing[, colSums(is.na(testing)) == 0])[8:59]
  head(testVariables)
   

# remove irrelevant variables 
irrelevantVar <- names(training.clean) %in% c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2",
                                         "cvtd_timestamp") 
training.clean <- training.clean[!irrelevantVar]

## after irrelevant variables removed
dim(training.clean)

# discard NA's which are more than 95%
iNA <- sapply(training.clean, is.na)
sumOfNA <- colSums(iNA)

NAPercent <- sumOfNA/(dim(training.clean)[1])
 
training.clean <- training.clean[,!(NAPercent > 0.95)]

## after removing NA's which are more than 95%
dim(training.clean) 

```

## Partition Data

```{r partation data}
set.seed(100)
nLen <- length(training.clean)


inTraining <- createDataPartition(training.clean$classe, p = .6)[[1]]
training.clean <- training.clean[inTraining,]; testing.clean <- training.clean[-inTraining,]

dim(training.clean)

```

```{r create model}

### training control setting
tc <- trainControl(method = "cv", number = 10)

### model fititng
set.seed(100)
fitModel <- train(classe ~., method="rpart", data=training.clean, trControl = tc)

results <- confusionMatrix(testing.clean$classe, predict(fitModel, newdata = testing.clean))

### model fitting after preprocess
fitModelAfter <- train(classe ~., method="rpart", preProcess = c("center", "scale"), data=training.clean,trControl = tc)

resultsAfter <- confusionMatrix(testing.clean$classe, predict(fitModelAfter, newdata = testing.clean))

### 1st model
results

### After PreProcessing model
resultsAfter

```

#### Even with PreProcessing, the Accuracy is around 0.5 its not convincing so not acceptable

#### Use Random Forest
 
```{r randomForest}
 
 data <- training.clean
 data$classe <- as.numeric(training.clean$classe)
 corr <- data.frame(cor(data[,-c(1)]))
 corr$name <- names(data[2:52])
 
 x <- data.frame(cbind(corr$classe, corr$name))
 names(x) <- c("cor","name")
  
 ### variables with high correlation with classe
tail(arrange(x, cor), 8)

### use variables wiht highest correlation 
fitModelHighCorr <- randomForest(classe ~ pitch_forearm + magnet_arm_x + accel_arm_x + total_accel_forearm + magnet_dumbbell_z + accel_dumbbell_x , data= training.clean)

rfResults <- confusionMatrix(testing.clean$classe, predict(fitModelHighCorr, newdata = testing.clean, type = "class"))

### shpw result
rfResults

### Full Model
fitModelFullBlown <- randomForest(classe ~., data= training.clean)
resultsFullBlown <- confusionMatrix(testing.clean$classe, predict(fitModelFullBlown, newdata = testing.clean, type = "class"))

### show full results
resultsFullBlown
```

# Cross Validation
```{r crossvalidation}
set.seed =100

dim(training.clean) 


splits <- split(training.clean, f =rep_len(1:10, nrow(training.clean) ))


### fix 10 equal data size
 allData <- function(index) {
   df <- splits[[ index[1] ]]
   for (i in 2:length(index)) df <- rbind(df, splits[[ index[i] ]])
    df
 } 



 cvResults <- as.data.frame(matrix(nrow = 7, ncol = 10))
 dim(cvResults)
 
index <- 1:10
 

for(var in 1:10){
 cvData <- allData(index[index!= var])
  cvFitModel <- randomForest(classe ~., data = cvData)
   cvFitModelResults <- confusionMatrix(splits[[var]]$classe, predict(cvFitModel, type = "class", newdata = splits[[var]]))
  
  cvResults[,var] <-cvFitModelResults$overall
}

#head(cvFitModelResults)
```

